Geometrization of quasicharacters of algebraic tori over local fields

### 2014-01-22 dictionary.tex
Minor improvements to both.

In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notation of {\it quasicharacter sheaves on $G$}.

A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions. The main result of this paper is the following statement (see Theorem~\ref{thm:snake}, which says more): 
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and that the trace of Frobenius morphism provides an isomorphism
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx).
\end{equation}
In this way we find the geometrization and categorification of characters of $G(\Fq)$.

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to disconnected algebraic groups over finite fields. Consider the etale group scheme $G = \ZZ_{\Fq}$ over $\Fq$; this is a $\QC$-scheme.
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general; for example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is not finite but bounded, which the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

@@

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

>>>>

In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions. The main result of this paper is the following statement (see Theorem~\ref{thm:snake}, which says more): 
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and the trace of Frobenius morphism provides an isomorphism
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx).
\end{equation}
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to disconnected algebraic groups over finite fields. Consider the etale group scheme $G = \ZZ_{\Fq}$ over $\Fq$; this is a $\QC$-scheme.
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. For example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is bounded but not finite, while the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

@@

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.  These subcategories play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

### 2014-01-22 geometrization.tex
Minor improvements to both.

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes over finite type over fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 As an application, we categorify a transfer of certain characters between characteristic $p$ tori and characteristic $0$ tori.
  \end{abstract}

@@

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:GN} we review the Greenberg transform of N\'eron models of tori, which produces
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses Theorem~\ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar}, \ref{sec:quasichar} and \ref{sec:wrK} also include
functoriality results that we hope to use in a future discussion of 
local class field theory.
In Section~\ref{sec:transfer} we discuss a transfer principle which allows
a geometric transfer of characters between characteristic $p$ tori and characteristic $0$ tori.
We close the paper with Section~\ref{sec:reciprocity}, in which we suggest that quasicharacter
sheaves may play a key role in a geometric and categorical form of the local Langlands
correspondence for algebraic tori over local fields.

>>>>

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes locally of finite type over finite fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 As an application, we categorify a transfer of certain characters between characteristic $p$ tori and characteristic $0$ tori.
  \end{abstract}

@@

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:GN} we review the Greenberg transform of N\'eron models of tori, which produces
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses \cite{cunningham-roe:dictionary} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar}, \ref{sec:quasichar} and \ref{sec:wrK} also include
functoriality results that we hope to use in a future discussion of 
local class field theory.
In Section~\ref{sec:transfer} we discuss a transfer principle which allows
a geometric transfer of characters between characteristic $p$ tori and characteristic $0$ tori.
We close the paper with Section~\ref{sec:reciprocity}, in which we suggest that quasicharacter
sheaves may play a key role in a geometric and categorical form of the local Langlands
correspondence for algebraic tori over local fields.
